Creating figures for the paper

### 2024-08-20 infercnv/reports/paper_figures.Rmd
using chatgpt to write new lolliplot function

# libraries
library(magrittr)
library(ggplot2)
library(SummarizedExperiment)
library(SingleCellExperiment)

# chunk options
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(warning = FALSE, dpi = 300, message = FALSE)
```

# Datasets

## LCM NanoSeq

>>>>

# libraries and functions
source("bin/lolliplot.R")
library(magrittr)
library(ggplot2)
library(SummarizedExperiment)
library(SingleCellExperiment)

# dirs
lh22_dir <- "/lustre/scratch125/casm/team268im/lh22/Rheumatoid_Arthritis_Paper/"
nb15_dir <- "/lustre/scratch126/casm/team154pc/nb15/liver/"

# chunk options
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(warning = FALSE, dpi = 300, message = FALSE)

# plotting params
filterNonCoding <- TRUE
pointWidth <- 0.017
mut_impact_palette <-
  c("Non-coding" = "gray100",
    "Synonymous" = "gray76",
    "Missense" = "#abd9e9",
    "Nonsense" = "#fdae61",
    "Indel" = "pink",
    "Essential_Splice" = "#fee090",
    "Stop_loss" = "#bb0f12")
```

First, we load the NanoSeq output and run dNdScv to annotate the mutations and 
to get a list of significantly positively selected genes.

```{r load_nanoseq_muts}
# load the annotated mutations from nanoseq output (`nano_muts`)
load(file.path(lh22_dir, "Figures/Figure_2/LCM_NanoSeq_Overlap/GOI",
               "pileup_muts.rda"))
nano_muts <- nano_muts %>% tibble::as_tibble()

# run dNdScv
dndscv_in <- nano_muts %>% dplyr::select(sampleID, chr, pos, ref, mut)
dndscv_out <- dndscv::dndscv(dndscv_in)

# combine annotated mutations with dnds output
nanoseq_data <-
  nano_muts %>%
  dplyr::mutate(
    impact = dplyr::case_when(impact == "no-SNV" ~ "Indel",
                              is.na(impact) ~ "Non-coding",
                              TRUE ~ impact),
    name = dplyr::case_when(impact == "Indel" ~
      # names for indels
      ntchange %>%
        tidyr::replace_na(".") %>%
        stringr::str_remove(".*-") %>%
        stringr::str_replace("frshift", "FS") %>%
        stringr::str_replace("inframe", "IF"),
      # names for snvs
      TRUE ~ tidyr::replace_na(aachange, "."))) %>%
  dplyr::select(sampleID, chr, pos, ref, mut, duplex_cov, impact, donor, name) %>%
  dplyr::left_join(dndscv_out$annotmuts) %>%
  # rename cols
  dplyr::rename(CHROM = chr, POS = pos, REF = ref, ALT = mut)
```

Now, we get coding sequence annotations for these genes of interest, including 
the TSS, the bounds of exons, and the protein domains.

```{r get_representative_domains, eval = F}
# TODO: dplyr::filter(protein_representative == TRUE)
accession <- "P21359"
url <-
  paste0(
    "https://www.ebi.ac.uk/interpro/api/entry/all/protein/UniProt/",
    accession,
    "/?extra_fields=hierarchy%2Cshort_name&amp;format=tsv&amp;page_size=200")
response <- httr::GET(url)
content <- httr::content(response)
domains <-
  content$results %>%
  purrr::map(function(domain) {
    md <-
      domain$metadata %>%
      purrr::compact() %>%
      tibble::as_tibble() %>%
      dplyr::rename_with(~ paste0("metadata_", .), .cols = everything())
    prot <-
      domain$proteins[[1]] %>%
      purrr::compact() %>%
      tibble::as_tibble() %>%
      tidyr::unnest_wider(entry_protein_locations) %>%
      dplyr::mutate(fragments = fragments[[1]]) %>%
      tidyr::unnest_wider(fragments, names_sep = "_") %>%
      dplyr::rename_with(~ paste0("protein_", .), .cols = everything())
    # extra <-
    #   domain$extra_fields %>%
    #   purrr::compact() %>%
    #   tibble::as_tibble() %>%
    #   tidyr::unnest_wider(hierarchy, names_sep = "_")
    dplyr::bind_cols(md, prot)
  }) %>%
  dplyr::bind_rows()
```

```{r get_gene_info}
# load refcds, add gene names
data("refcds_hg19", package = "dndscv")
names(RefCDS) <- RefCDS %>% purrr::map(~ .$gene_name)

#  get gene info
goi <-
  list("NF1", "TNFRSF14") %>%
  purrr::set_names() %>%
  purrr::map(function(gene) {

    # get gene info
    g <- RefCDS[[gene]]

    # get cds intervals
    intervals_cds0 <- tibble::tibble(
      chr = g$chr,
      strand = g$strand,
      start = g$intervals_cds[, 1],
      end = g$intervals_cds[, 2]
    ) %>%
      dplyr::arrange(start) %>%
      dplyr::mutate(length = (end + 1) - start)

    # get genomic position to CDS position index
    index <- getCDSindex(intervals_cds0)

    # convert CDS intervals
    intervals_cds <-
      intervals_cds0 %>%
      dplyr::left_join(dplyr::select(index, start = pos, cdsStart = cdsPos),
                       by = "start") %>%
      dplyr::left_join(dplyr::select(index, end = pos, cdsEnd = cdsPos),
                       by = "end")

    # get protein domains from InterPro REST API
    domains <- getProteinDomains(gene)

    # subset nanoseq_data to variants in the gene
    gene_cds_muts <-
      nanoseq_data %>%
      dplyr::filter(CHROM == unique(intervals_cds$chr),
                    POS >= min(intervals_cds$start),
                    POS <= max(intervals_cds$end))

    # get coords outside the CDS
    gene_non_cds_muts <-
      index %>%
      dplyr::rename(POS = pos) %>%
      dplyr::bind_rows(getCDSindexNonCDS(index, gene_cds_muts,
                                         strand = unique(intervals_cds$strand)))

    # get all muts in the gene
    gene_muts <-
      gene_cds_muts %>%
      dplyr::left_join(gene_non_cds_muts, by = "POS") %>%
      dplyr::arrange(cdsPos) %>%
      tidyr::drop_na(cdsPos) %>%
      # count number of patients with the mutation
      dplyr::group_by(CHROM, POS, REF, ALT) %>%
      dplyr::mutate(n_donors = dplyr::n_distinct(pid)) %>%
      dplyr::ungroup()

    # get label positions
    lab_pos <-
      jitterLabels(sort(gene_muts$cdsPos),
                   xscale = c(1, max(intervals_cds$cdsEnd)),
                   lineW = pointWidth) %>%
      reAdjustLabels(lineW = pointWidth, xmax = max(intervals_cds$cdsEnd))

    # prep data for plotting
    p_dat <-
      gene_muts %>%
      dplyr::mutate(lab_pos = lab_pos,
                    impact = factor(impact, levels = names(mut_impact_palette)),
                    aachange = tidyr::replace_na(aachange, "."))

    # get plotting limits
    x_max <-
      max(c(intervals_cds$cdsEnd, intervals_cds$cdsStart)) +
      max(c(intervals_cds$cdsEnd, intervals_cds$cdsStart)) * 0.02
    x_min <- 1 - x_max * 0.01
    if (min(p_dat$lab_pos < x_min)) x_min <- 1 - x_max * 0.02
    legend_pos <- c(0.5, 0.95)
    y_max_count <- max(p_dat$n_donors) + 2
    if (y_max_count > 10) y_max_count <- y_max_count + 2

    # plot
    p <-
      ggplot(p_dat, aes(x = cdsPos, y = n_donors)) +
      # variant stems
      geom_segment(aes(x = cdsPos, xend = cdsPos, y = 0, yend = (0.2 * n_donors))) +
      geom_segment(aes(x = cdsPos, xend = lab_pos, y = 0.2 * n_donors, yend = 0.3 * n_donors)) +
      geom_segment(aes(x = lab_pos, xend = lab_pos, y = 0.3 * n_donors, yend = n_donors)) +
      #Variant points
      geom_point(aes(x = lab_pos, y = n_donors, fill=impact), shape=21,  color="black", size=5) +
      #Label aa change
      geom_text(aes(label = name, x = lab_pos, y = (n_donors + y_max_count/8)), angle = 90) + 
      ggtitle(gene) +
      # label clinvar
      # geom_text(aes(label = label, x = lab_pos, y = (n_donors)), size = 3.5) +
      scale_fill_manual(guide = "legend", values = mut_impact_palette) +
      scale_x_continuous(expand = c(0, 0), limits = c(x_min, x_max)) +
      ylab("Sperm samples with mutation") +
      coord_cartesian(clip = "off") +
      scale_y_continuous(limits = c(0, y_max_count), expand = c(0, 0),
                         sec.axis = sec_axis(~ . * (200000/y_max_count),
                         name = "Duplex Coverage")) +
      theme(
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA),
        panel.grid = element_blank(),
        panel.border = element_blank(),
        plot.margin = unit(c(0.5, 0.5, 0, 0.5), "cm"),
        panel.spacing = unit(c(0, 0, 0, 0), "cm"),
        axis.text.y = element_text(margin = margin(0,0.3,0,0.2, "cm")),
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.line = element_blank(),
        plot.title = element_text(hjust = 0.5),
        axis.ticks.length.x = unit(0, "null"),
        legend.position = legend_pos,
        legend.direction = "horizontal",
        legend.title = element_blank(),
        legend.key = element_rect(fill = NA))

    # plot mutations
    p_muts <-
      p_dat %>%
      dplyr::group_by(CHROM, POS, REF, ALT) %>%
      dplyr::mutate(n_donors = dplyr::n_distinct(donor)) %>%
      ggplot() +
      # variant stems
      geom_segment(aes(x = cdsPos, xend = cdsPos, y = 0,
                       yend = (0.125 * max(lab_pos)))) +
      geom_segment(aes(x = cdsPos, xend = labPos, y = (0.125 * max(lab_pos)),
                       yend = (0.25 * max(lab_pos)))) +
      geom_segment(aes(x = labPos, xend = labPos, y = (0.25 * max(lab_pos)),
                       yend = (0.325 * max(lab_pos)))) +
      # variant points
      geom_point(aes(x = labPos, y = 0.325 * max(lab_pos), fill = impact,
                     size = n_donors),
                shape = 21, colour = "black", size = 5) +
      # label aa change
      # geom_text(aes(label = name, x = labPos, y = (0.5 * max(lab_pos))),
      #           angle = 90) +
      theme_void() +
      scale_fill_manual(guide = "legend", values = mut_impact_palette) +
      scale_x_continuous(expand = c(0, 0), limits = c(x_min, x_max)) +
      coord_cartesian(clip = "off") +
      scale_y_continuous(limits = c(0, max(lab_pos)), expand = c(0, 0)) +
      ggtitle(g)

    # plot CDS
    p_cds <-
      intervals_cds %>%
      ggplot(aes(xmin = cdsStart, xmax = cdsEnd, ymin = 0, ymax = 1)) +
      geom_rect(fill = "#d0d1e6", colour = "black") +
      theme_void() + 
      xlab("CDS") +
      scale_x_continuous(expand = c(0, 0), limits = c(x_min, x_max)) +
      scale_y_continuous(expand = c(0, 0)) +
      theme(plot.margin = unit(c(0, 0.5, 0.5, 0.5), "cm"))

    pdf("test.pdf", width = 20)
    print(p)
    cowplot::plot_grid(p_muts, p_cds, ncol = 1, rel_heights = c(15, 1))
    dev.off()

    # return all
    list(gene_muts = gene_muts,
         intervals_cds = intervals_cds,
         domains = domains,
         index = index)
  })
```

```{r plot_lolliplot}
# set label positions
lab_pos <-
  jitterLabels(sort(gene$cdsPos),
               xscale = c(1, max(cds_coords_index$cdsPos)),
               lineW = pointWidth) %>%
  reAdjustLabels(lineW = pointWidth, xmax = max(cds_coords_index$cdsPos))
variants <-
  gene %>%
  dplyr::mutate(labPos = lab_pos,
                aachange = tidyr::replace_na(aachange, "."))

# xmax and xmin
x_max <-
  max(c(intervals_cds$cdsEnd, intervals_cds$cdsStart)) +
  max(c(intervals_cds$cdsEnd, intervals_cds$cdsStart)) * 0.05
x_min <- 1 - x_max * 0.01

# plot variants
p_var <-
  variants %>%
  dplyr::group_by(CHROM, POS, REF, ALT) %>%
  dplyr::mutate(n_donors = dplyr::n_distinct(donor)) %>%
  ggplot() +
  # variant stems
  geom_segment(aes(x = cdsPos, xend = cdsPos, y = 0,
                   yend = (0.125 * max(lab_pos)))) +
  geom_segment(aes(x = cdsPos, xend = labPos, y = (0.125 * max(lab_pos)),
                   yend = (0.25 * max(lab_pos)))) +
  geom_segment(aes(x = labPos, xend = labPos, y = (0.25 * max(lab_pos)),
                   yend = (0.325 * max(lab_pos)))) +
  # variant points
  geom_point(aes(x = labPos, y = 0.325 * max(lab_pos), fill = impact,
                 size = n_donors),
             shape = 21, colour = "black", size = 5) +
  # label aa change
  # geom_text(aes(label = name, x = labPos, y = (0.5 * max(lab_pos))),
  #           angle = 90) +
  theme_void() +
  scale_fill_manual(guide = "legend", values = mut_impact_palette) +
  scale_x_continuous(expand = c(0, 0), limits = c(x_min, x_max)) +
  coord_cartesian(clip = "off") +
  scale_y_continuous(limits = c(0, max(lab_pos)), expand = c(0, 0)) +
  ggtitle(g)

# plot CDS
p_cds <-
  intervals_cds %>%
  ggplot(aes(xmin = cdsStart, xmax = cdsEnd, ymin = 0, ymax = 1)) +
  geom_rect(fill = "#d0d1e6", colour = "black") +
  theme_void() + 
  xlab("CDS") +
  scale_x_continuous(expand = c(0, 0), limits = c(x_min, x_max)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme(plot.margin = unit(c(0, 0.5, 0.5, 0.5), "cm"))

pdf("test.pdf", width = 20)
cowplot::plot_grid(p_var, p_cds, ncol = 1, rel_heights = c(15, 1))
dev.off()

```
